Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-17 analysis/paper/paper.qmd
Edited until methods

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterizes the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations. Our results show that the stemmed points were could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different TCSA range of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were mainly produced before Last Glacial Maximum (LGM). We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern is confirmed. We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.

@@

The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods. The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points. We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these change over time. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses? As part of our discussion, we examine possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged]. Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did]. Therefore, as @o2018origins claimed, understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.

Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. Debate about explaining the origin of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (i.e. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade cores. While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages and found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points. Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

The connection between stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human]. Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late Marine isotope stages (MIS) 3. In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].

@@

Though stemmed points have been generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components. Preserved wooden components of projectile tools are too few and rare to standardize the overall scale and variability of the tools used in prehistory [@shea2006origins]. @Lee_Kong_2002 claimed stemmed points were generalized 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape. Other researchers propose that non-symmetrical stemmed points, with retouch on one side of the artifact or denticulate blades on one side, should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011]. There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim_2017]. Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002]. On Yongsandong site, for example, the percentage of complete tools is only 10%, while the percentage of missing tips is 33%. On Jingeuneul site, the percentages are 16% and 50%, respectively [@Park_2013]. In addition to the morphological aspect of stemmed points, investigation of a whole site and tool composition of an assemblage shows that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].

@@

Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized]. It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and validated by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.

A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth, while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons because they have not been preserved in the archaeological record [@lombard2020tip]. @sisk2011african also mentioned that TCSP cannot be applied to backed pieces that were hafted as projectile armatures.

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site. For TCSA, they were only able to use 10 stemmed points because they chose to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the range of TCSA values for Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft. They conclude that stemmed points may have been used as spear-throwers or bows. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for a much larger sample of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges). Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses?

>>>>

abstract: |
 The transition from the Early to Late Paleolithic in Korea is characterized by the introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials. Stemmed points have been considered a representative tool that led this set of changes. In this study, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic period. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations. Our results show that the stemmed points could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different TCSA ranges of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were predominantly produced before the Last Glacial Maximum (LGM). We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern was confirmed. We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.

@@

The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc. [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

The purpose of this study is to examine the possible uses of stemmed points in order to understand how stemmed points played a role in the technological transition. We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these change over time. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses? As part of our discussion, we examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer referring to them as 'stemmed points' here to distinguish them from Bronze Age stone projectile points which have long been called 'tanged points' in Korea [@park2022did]. Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.

Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, a stone point and a shaft, presumably made out of wood [@seong2008tanged]. Using blades as the blank for the point enables mass production of this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did]. Therefore, as @o2018origins claimed, understanding the emergence of projectile technology provides insights into greater cultural, evolutionary, and behavioral cognitive flexibility.

Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. Debate about explaining the origin of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (i.e. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how they explain a few early sites that contain stemmed points but no blades and blade cores. While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages. The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points. Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

The connection between stemmed points in both Korea and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human]. Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca. 41.5 ka, and made out of elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appeared in Kyushu, Japan during the late Marine isotope stage (MIS) 3. In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki  in Japanese), backed knives, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].

@@

Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to determine their likely uses without knowing their complete shape when attached to other components. Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins]. @Lee_Kong_2002 claimed stemmed points should be considered more generally as 'stemmed tools' because of their diverse shapes, which could represent different functions, and the uncertainty of their complete shape. Other researchers propose that non-symmetrical stemmed points, with retouch on one side or denticulate blades on one side, should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].

Stemmed points are typically symmetrical from tip to tang, with the central axis serving as a line of symmetry [@Lee_Jang_2011]. There is a high percentage of broken tips and stems, and the reused tools were repaired in accordance with symmetry [@Park_2013; @Kim_2017]. Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002]. For example, at Yongsandong site, only 10% of the tools are complete, while 33% of the tips are missing. In the case of Jingeuneul, the percentages are 16% and 50%, respectively [Park_2013]. In addition to the morphological aspect of stemmed points, investigations of a whole site and the tool composition of an assemblage suggest that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].

@@

Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized]. It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.

One of the key advantages of the TCSA metric is its convenience of application: regardless of the point type, only the maximum width and thickness measurements are required to calculate the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for a more precise measure of the force required to penetrate a target to a lethal depth, whereas the TCSA metric is more associated with cutting. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons because they have not been preserved in the archaeological record [@lombard2020tip]. @sisk2011african also mentioned that TCSP cannot be applied to backed pieces that were hafted as projectile armatures.

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. They analyzed stemmed points from Jingeuneul, located in the southwest of Korea, which has the largest number of stemmed points (n = 99) found at a single site to date. For the TCSA, they were only able to use ten stemmed points since they selected stemmed points that retained the widest and thickest parts of the specimens and showed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the range of TCSA values for Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft. Based on the results, they conclude that stemmed points may have been used as spear-throwers or bows. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for a much larger sample of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.

@@

After the first discovery at the Seokjangri site at the 1960s, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges). Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses?
